---
title: "01_QC_integration_clustering"
author: "Divya"
date: "2026-08-14"
output: html_document
---

##load required packages

```{r, results = FALSE, message=FALSE, warning=FALSE}
# Load required packages
library(dplyr) 
library(Seurat) 
library(patchwork) 
library(ggplot2) 
library(cowplot) 
library(rmarkdown) 
library(SummarizedExperiment) 
library(htmlwidgets) 
library(htmltools) 
library(shiny) 
library(openxlsx) 
library(hdf5r) 
library(harmony) 
library(DT)
library(SeuratWrappers)

```


## Assign samples

```{r, results = FALSE, message=FALSE, warning=FALSE}
sampleNames <- c("IBD1","IBD2","IBD3","IBD4","IBD5","IBD6","IBD7","IBD8","IBD9","IBD10","IBD11","IBD12","IBD13","IBD14","IBD15","NV1","NV2","NV3","NV4","NV5","NV6","NV7")
data_dir <- (".../IEL")
projectName <- "IEL"
h5_string <- ".h5"
```


```{r,  eval=FALSE, message = FALSE, results= FALSE, warning=FALSE}
# Define filenames and path
filenames <- list.files(path = data_dir, pattern = "\\.h5$", full.names = TRUE)

# Create a list to store Seurat objects
seurat_list <- list()

# Loop through filenames and create Seurat objects
for (i in 1:length(filenames)) {
  raw_data <- Read10X_h5(filenames[i], use.names = TRUE, unique.features = TRUE)
  seurat_obj <- CreateSeuratObject(counts = raw_data,
                                    project = projectName,
                                   min.cells = 3,
                                   min.features = 200)
# Add sample name to orig.ident
 seurat_obj$orig.ident <- sampleNames[i]

# Store in the list
 seurat_list[[i]] <- seurat_obj
}
 
# Merge all Seurat objects into one
combined_data <- merge(
  x = seurat_list[[1]], 
  y = seurat_list[-1], 
  add.cell.ids = sampleNames)
 
combined_data$sample_names <- combined_data$orig.ident

```

# Add conditions

```{r, eval= FALSE, results = FALSE, message=FALSE, warning=FALSE}
sample_to_group <- setNames(
  ifelse(grepl("NV", sampleNames), "Healthy",
         ifelse(grepl("IBD|UC", sampleNames), "UC", NA)),
  sampleNames
)
# Apply the new annotations to your Seurat object
combined_data$condition <- plyr::mapvalues(
  combined_data$sample_names,
   from = names(sample_to_group),
  to = sample_to_group
)

# Check if condition has been mapped appropriately
unique(combined_data$condition)
unique(combined_data$sample_names)

#save the raw file
saveRDS(combined_data, "../file_raw.RDS")
```

##

```{r, eval = FALSE, results = FALSE, message=FALSE, warning=FALSE}
# Standard Seurat workflow
data <- readRDS("../file_raw.RDS")

# Standard normalisation workflow
data <- NormalizeData(data, normalization.method = "LogNormalize", scale.factor = 10000)
data <- FindVariableFeatures(data, selection.method = "vst", nfeatures = 2000)
data <- ScaleData(data, verbose = TRUE)

#dimension reduction
data <- RunPCA(data, npcs = 50, verbose = FALSE)

# Check how many PCs to use
ElbowPlot(data, ndims = 50)

# Use appropriate number of PCs
data <- RunUMAP(data, reduction = "pca", dims = 1:20)

##save raw_UMAP_file
saveRDS(data, "../raw_UMAP.RDS")
```

# Non-integrated Samples before QC 

```{r, results = 'asis', message=FALSE, warning=FALSE}
data <- readRDS("../raw_UMAP.RDS")
cat("##", "Samples", " \n")
DimPlot(data, reduction = "umap", group.by = "sample_names", raster = FALSE) + ggtitle ("Samples before integration and QC") + theme(legend.position = "right")
cat(" \n \n")
cat("##", "Condition", " \n")
DimPlot(data, reduction = "umap", group.by = "condition", raster = FALSE) +
  ggtitle("Condition before integration and QC") +
  theme(legend.position = "right")
cat(" \n \n")

```


# QC: Checking mitochondrial and ribosomal gene percentages

```{r, results = FALSE, message=FALSE, warning=FALSE}
# Check mitochondrial gene names
gene_names <- rownames(data)
mt_genes <- grep("^MT-", gene_names, value = TRUE)
print(mt_genes)

# Calculate mitochondrial gene percentage
data[["percent.mt"]] <- PercentageFeatureSet(data, pattern = "^MT-")
 
# Add new column with mitochondrial percentage information
data$mt_label <- factor(ifelse(data$percent.mt > 20, "Above 20%", "Below 20%"))

# Check ribosomal gene names
gene_names <- rownames(data)
rp_genes <- grep("^RP[SL]", gene_names, value = TRUE)
print(rp_genes)

# Calculate ribosomal gene percentage
data[["percent.rp"]] <- PercentageFeatureSet(data, pattern = "^RP[SL]")

# Add new column with ribosomal percentage information
data$rp_label <- factor(ifelse(data$percent.rp > 10, "Above 10%", "Below 10%"))

# Save RDS with mitochondrial and ribosomal percentages
saveRDS(data, file = paste("/Users/divyajha/Desktop/R_analysis/GDT/data_combined_raw_umap.RDS"))
```

# Plot and save integrated mitochondrial and ribosomal pct
```{r, results = 'asis', message=FALSE, warning=FALSE}
data <- readRDS(".../raw_UMAP.RDS")
cat("##", "Mitochondrial %", " \n")
DimPlot(data, reduction = "umap", group.by = "mt_label", raster = FALSE) + ggtitle ("Percentage of Mitochondrial Genes")
VlnPlot(data, features = c("percent.mt"), pt.size = 0.0, group.by = "sample_names") + ggtitle ("Percentage of Mitochondrial Genes") + theme(legend.position = "none") + coord_cartesian(ylim = c(0, 25))
cat(" \n \n")

cat("##", "Ribosomal %", " \n")
DimPlot(data, reduction = "umap", group.by = "rp_label", raster = FALSE) + ggtitle ("Percentage of Ribosomal Genes")
VlnPlot(data, features = c("percent.rp"), pt.size = 0.0, group.by = "sample_names") + ggtitle ("Percentage of Ribosomal Genes") + theme(legend.position = "none") + coord_cartesian(ylim = c(0, 25))
cat(" \n \n")

pdf(paste("../pdfs/non_integrated_mitochondrial_pct.pdf"), width=16, height=10, useDingbats = FALSE)
DimPlot(data, reduction = "umap", group.by = "mt_label", raster = FALSE) + ggtitle ("Percentage of Mitochondrial Genes")
VlnPlot(data, features = c("percent.mt"), pt.size = 0.0, group.by = "sample") + ggtitle ("Percentage of Mitochondrial Genes") + theme(legend.position = "none") + coord_cartesian(ylim = c(0, 25))
dev.off()
```

# Filter out cells with >20% mitochondrial gene exp and nFeature > 200 or < 2500 and save
```{r, results = FALSE, message=FALSE, warning=FALSE}
data_filtered <- subset(data, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 20)

# Save RDS post filtering
saveRDS(data_filtered, file = paste("../data_filtered_umap.RDS"))
```

# Read pre and post filtering RDS files to compute statistics on how many cells are being filtered

```{r, results=FALSE, message=FALSE, warning=FALSE}
data_filtered <- readRDS("../data_filtered_umap.RDS")
data <- readRDS("../raw_UMAP.RDS")

# Get total and filtered cell counts 
total_cells <- length(Cells(data))
retained_cells <- length(Cells(data_filtered))
filtered_cells <- total_cells - retained_cells  

# Calculate percentages
retained_percentage <- (retained_cells / total_cells) * 100
filtered_percentage <- (filtered_cells / total_cells) * 100

# Count cells with NA values in filtering columns
na_filtered <- sum(is.na(data$nFeature_RNA) | is.na(data$percent.mt))

# Count cells removed due to BOTH filters (overlap)
overlap_filtered <- sum((data$nFeature_RNA < 200 | data$nFeature_RNA > 2500) & (data$percent.mt > 20), na.rm = TRUE)

# Count cells removed for nFeature_RNA and percent.mt, adjusting for overlap
nfeature_count <- sum(data$nFeature_RNA < 200 | data$nFeature_RNA > 2500, na.rm = TRUE) 
mitochondrial_cells <- sum(data$percent.mt > 20, na.rm = TRUE)

# Correct the total filtered count by adding NAs
filtered_cells_corrected <- nfeature_count + mitochondrial_cells - overlap_filtered + na_filtered

# Summary table
summary_table <- data.frame(
  "Category" = c("Total Cells", "Retained Cells", "Total Filtered Cells", "nFeature_RNA Cells", "Mitochondrial Cells"),
  "Raw Count" = c(total_cells, retained_cells, filtered_cells, nfeature_count, mitochondrial_cells),
  "Percentage" = c(
    100, 
    retained_percentage, 
    (filtered_cells / total_cells) * 100, 
    (nfeature_count / total_cells) * 100, 
    (mitochondrial_cells / total_cells) * 100
  )
)

# Format to remove scientific notation
summary_table$Percentage <- format(summary_table$Percentage, scientific = FALSE)
```

# Cell Count Summary

```{r, results='asis', message=FALSE, warning=FALSE}
# Display summary table
paged_table(summary_table)
```


# Non-integrated Samples after mitochondrial and nFeature RNA Filtering
```{r, results = 'asis', echo=FALSE, message=FALSE, warning=FALSE}

data_filtered <- readRDS(".../data_filtered_umap.RDS")
cat("##", "Samples", " \n")
DimPlot(data, reduction = "umap", group.by = "sample_names", raster = FALSE) + ggtitle ("UMAP post Mitochondrial Filtering") + theme(legend.position = "right")

cat(" \n \n")
cat("##", "Condition", " \n")
DimPlot(data, reduction = "umap", group.by = "condition", raster = FALSE) + ggtitle ("UMAP post Mitochondrial Filtering")
cat(" \n \n")
```


## Calculate cell distribution sample wise, raw counts and percentages

```{r, results=FALSE, message=FALSE, warning=FALSE}
# Calculate cell distribution sample wise, raw counts and percentages

# Load filtered dataset
# data_filtered <- readRDS("../data_filtered_umap.RDS")

# Count number of cells per sample
cell_counts <- table(data_filtered$sample_names)

# Convert to data frame
df_counts <- as.data.frame(cell_counts)
colnames(df_counts) <- c("Sample", "Cell_Count")

# Calculate percentage distribution (ensuring it sums to 100)
df_counts$Percentage <- (df_counts$Cell_Count / sum(df_counts$Cell_Count)) * 100

# Bar plot for percentage distribution with y-axis fixed from 0 to 100
p1 <- ggplot(df_counts, aes(x = Sample, y = Percentage, fill = Sample)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Cell Percentage Distribution Across Samples",
    x = "Sample",
    y = "Percentage of Total Cells",
    fill = "Sample"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

# Bar plot for raw counts
p2 <- ggplot(df_counts, aes(x = Sample, y = Cell_Count, fill = Sample)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Raw Cell Count Distribution Across Samples",
    x = "Sample",
    y = "Number of Cells",
    fill = "Sample"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

cat("##", "Raw Count Distribution", " \n")
print(p2)
cat ('\n \n')

cat("##", "Percentage Distribution", " \n")
print(p1)
cat ('\n \n')
```

# Integration of data to remove artifacts/batch effect

```{r,  eval=FALSE, message = FALSE, results= FALSE, warning=FALSE}
data.combined <- readRDS("../data_filtered_umap.RDS")
data.combined <- RunHarmony(object = data.combined, group.by.vars = "sample_names")
# Save data
saveRDS(data.combined, file = paste("../data_Integrated.RDS"))

# Run UMAP using the Harmony embeddings, not the original PCA
data.combined <- RunUMAP(data.combined, reduction = "harmony", dims = 1:20)
# Save integrated data
saveRDS(data.combined, file = paste("../data_Integrated_umap.RDS") )
```


# UMAP for Samples before integration

```{r,results='asis', message=FALSE, warning=FALSE}
# Load non-integrated data
data.bf <- readRDS("../data_filtered_umap.RDS")

DimPlot(data.bf, reduction = "umap", group.by = 'sample_names') + 
  ggtitle('Samples before integration') + 
  theme(legend.position = "right")

# Sample wise dimplot
sample_names <- unique(data.bf$sample_names)

for (samp in sample_names) {
  cat('##', samp, ' \n')
  print(
    DimPlot(
      subset(data.bf, subset = sample_names == samp),
      reduction = "umap",
      label = TRUE
    ) + ggtitle(samp)
  )
  cat(' \n \n')
}

```

# UMAP for Samples after integration

```{r,results='asis', message=FALSE, warning=FALSE}
# Load integrated data
# data.combined <- readRDS("../data_Integrated_umap.RDS")

## by sample names
cat('##','All samples',' \n')
DimPlot(data.combined, reduction = "umap", group.by= 'sample_names' ) + ggtitle('Samples after integration') + theme(legend.position = "right") 
cat(' \n \n')
## by condition
DimPlot(data.combined, reduction = "umap", group.by= 'condition' ) + ggtitle('Samples after integration') + theme(legend.position = "right") 
##extract the list of unique sample names
sample_names_list <- unique(data.combined$sample_names)

# Sample wise dimplot
for (samp in sample_names_list) {
  cat('##', samp, ' \n')
  print(
    DimPlot(
      subset(data.combined, subset = sample_names %in% c(samp)),
      reduction = "umap"
    ) + ggtitle(samp)
  )
  cat(' \n \n')
}
```


###Clustering for primary clusters

```{r, eval = FALSE, message = FALSE, results= FALSE, warning=FALSE}
data.combined <- FindNeighbors(data.combined, reduction = "harmony", dims = 1:20)
data.combined <- FindClusters(data.combined, resolution = 0.3)

saveRDS(data.combined, file = paste("../primary_clusters.RDS") )
```


##find markers for primary clusters

```{r, eval = FALSE, message=FALSE, warning=FALSE}
data.combined <- readRDS("../primary_clusters.RDS")
DefaultAssay(data.combined) <- "RNA"
Idents(data.combined) <- "seurat_clusters"
data.combined <- JoinLayers(data.combined)
data.markers <- FindAllMarkers(data.combined, only.pos = FALSE, min.pct = 0.25, logfc.threshold = 0.25)
# # Save markers
write.csv(data.markers, file='../markers_primary_clusters.csv')
```

###primary cell type annotations based on canonical markers

```{r, results = FALSE, message=FALSE, warning=FALSE}
data_reannotated <- readRDS("../primary_clusters.RDS")
Idents(data_reannotated) <- "seurat_clusters"

# Annotate all clusters appropriately
cluster_annotations <- c(  "0" = "Myeloid",
  "1" = "GDT cells",
  "2" = "ABT cells",
  "3" = "B cells",
  "4" = "Epi cells",
  "5" = "Plasma cells",
  "6" = "Cyc cells",
  "7" = "Myeloid",
  "8" = "Epi cells",
  "9" = "Epi cells",
  "10" = "Myeloid",
  "11" = "Epi cells",
  "12" = "GDT cells",
  "13" = "Mast cells",
  "14" = "ABT/ILC cells",
  "15" = "Stromal cells"
)

# Apply the new annotations 
data_reannotated$cluster_label <- plyr::mapvalues(
 Idents(data_reannotated),
    from = names(cluster_annotations),
  to = cluster_annotations
)

# Update the identities in the Seurat object
Idents(data_reannotated) <- data_reannotated$cluster_label

##save
saveRDS(data_reannotated, "../annotated_primary_clusters.RDS")

DimPlot(data_reannotated, reduction = "umap", label.size = 12, pt.size = 0.2, repel = TRUE+ theme(legend.position = "right"),raster = FALSE)

```


