---
title: "02_Subclustering_GDIEL"
author: "Divya"
date: "2026-08-14"
output: html_document
---

##load required packages

```{r, results = FALSE, message=FALSE, warning=FALSE}
# Load required packages
library(dplyr) 
library(Seurat) 
library(patchwork) 
library(ggplot2) 
library(cowplot) 
library(rmarkdown) 
library(SummarizedExperiment) 
library(htmlwidgets) 
library(htmltools) 
library(shiny) 
library(openxlsx) 
library(hdf5r) 
library(harmony) 
library(DT)
library(SeuratWrappers)

```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Subsetting data, scaling and integrating the subsetted data


```{r, results = FALSE, eval= FALSE, message=FALSE, warning=FALSE}
data.combined <- readRDS(".../annotated_primary_clusters.RDS")
Idents(data.combined) <- "seurat_clusters"
subset_data <- subset(data.combined, idents = c("1","12"))
subset_data <- FindVariableFeatures(subset_data, selection.method = "vst", nfeatures = 2000)
subset_data <- ScaleData(subset_data, verbose = FALSE)
 
#dimension reduction
subset_data <- RunPCA(subset_data, npcs = 50, verbose = FALSE)
subset_data <- RunHarmony(subset_data, "sample_names")

# Check how many PCs to use
ElbowPlot(subset_data, ndims = 50, reduction = "harmony")
subset_data <- RunUMAP(subset_data, reduction = "harmony", dims = 1:25)
# Save integrated subset data
saveRDS(subset_data, file = paste(".../subset_umap.RDS"))
```


# Subset UMAP before clustering. Sometimes subsets need to be reintegrated. Check if this is necessary by using DimPlot with samples as well as condition

```{r,results='asis', message=FALSE, warning=FALSE}
subset_data <- readRDS(".../subset_umap.RDS")
DimPlot(subset_data, reduction = "umap") + ggtitle ("Subset before clustering") 

# Extract counts for subset data
data.combined <- readRDS(".../annotated_primary_clusters.RDS")
subset_data <- readRDS(".../subset_umap.RDS")

# Get the raw count of GDT cells
subset_count <- length(Cells(subset_data))
# Get the total number of cells in the dataset
total_cells <- length(Cells(data.combined))
# Calculate the percentage of GDT in the dataset
subset_percentage <- (subset_count / total_cells) * 100

# Create a summary table
summary_table <- data.frame(
  "Cell Type" = c("GDTCSC Cells", "Total Cells"),
  "Raw Count" = c(subset_count, total_cells),
  "Percentage" = c(subset_percentage, 100)
)

#########
# Extract metadata for the subset (GDTSC cells)
meta <- subset_data@meta.data
# Count number of GDTSC cells per condition
gdt_counts_per_condition <- meta %>%
  group_by(condition) %>%
  summarise(GDTSC_count = n())
print(gdt_counts_per_condition)

########
# Assuming you have your full dataset metadata
meta <- data.combined@meta.data
# Filter to keep only GDTSC cells, then count per sample
gdt_counts_per_sample <- meta %>%
  filter(sample_names %in% c("IBDUC-", "NV-")) %>%  # keep only those samples
  group_by(sample_names) %>%
  summarise(GDTSC_count = n())
print(gdt_counts_per_sample)
```


# Sub-clustering for Subset cells

```{r, eval = FALSE, message = FALSE, results= FALSE, warning=FALSE}
# Do clustering
subset_data <- FindClusters(subset_data, resolution = 0.3)
# Save clustering data
saveRDS(subset_data, file = paste(".../subset_clusters.RDS") )
```

###annotate the clusters of subset data and annotate subtypes

```{r, results = FALSE, eval = FALSE, message=FALSE, warning=FALSE}
subset_data <- readRDS(".../subset_clusters.RDS")
Idents(subset_data) <- "seurat_clusters"
# Annotate all clusters 
cluster_annotations <- c(
"0" = "Terminally_differentiated_GDT",
"1" = "Activated_stem-like_GDT",
"2" = "GzmKhi_MHCiihi_GDT",
"3" = "Quiescent_stem_like_GDT",
"4" = "GD17",
"5" = "Epimixed_GDT",
"6" = "Mature_cytotoxic_GDT",
"7" = "IFNghi_activated_GDT",
"8" = "IL32hi_GDT"
)

# Apply the new annotations to your Seurat object
subset_data$cluster_label <- plyr::mapvalues(
  subset_data$seurat_clusters,
  from = names(cluster_annotations),
   to = cluster_annotations
)

# Update the identities in the Seurat object
Idents(subset_data) <- subset_data$cluster_label
###SAVE RDS
saveRDS(subset_data, ".../annotated_subset_clusters.RDS")
# Clustering UMAP
DimPlot(subset_data, reduction = "umap", label = TRUE,label.size = 4, repel = TRUE, raster=FALSE) + ggtitle("GDT cell seurat_clusters")

```


